\textbf{Are regime changes always bad economics? Evidence from daily financial data}

### 2019-09-09 doc/paper.tex
Updates to theory and examinations of additional events
Added some additional theoretical discussion to address comments, as well as added analysis and visualization of the 1992 Venezuela coup attempt.

The image of a coup or popular uprising ousting an autocratic ruler and installing a democratic government is an evocative one. However, both recent events and prior research suggest that coups rarely lead to democracy. In the wake of the Arab Spring---in which rulers were deposed in four countries---only Tunisia ultimately transitioned to democracy. These results are in line with prior research that claims that in autocracies coups rarely lead to democratization, and in democracies they cause instability \citep{derpanopoulos2015coups, marinov2014coups, powell2011global, thyne2016coup, varol2011democratic}.

Of course, coups and other regime changes not only impact a country's political environment but its economy as well. For instance, investors, multinational firms, development agencies, and aid organizations rank political risk as a top consideration when making investment decisions in emerging markets.\footnote{In the 2013 Multilateral Investment Guarantee Agency (MIGA) \citet{wipr2013} report (the last report published), executives of multinational enterprises (MNEs) ranked political risk as the second most important constraint for foreign direct investment (FDI) in developing countries over the next three years (after macroeconomic instability). In \citet{wipr2011} and  \citet{wipr2012}, political risk was ranked the most important constraint for FDI in developing countries - even greater than macroeconomic instability. The types of political risk of most concern to investors in developing countries (ranked in order of importance) were 1) adverse regulatory changes, 2) breaches of contract, 3) transfers and convertibility restrictions, 4) civil disturbances, 5) non-honoring of government guarantees, 6) expropriation nationalization, 7) terrorism and 8) war.} Common rationales are that unstable governments act myopically and adopt inefficient policies \citep{svensson1998investment,devereux1998political}\footnote{\citet{svensson1998investment} argues that governments are less likely to invest in the legal system and the protection of property rights when political uncertainty is high. Similarly, \citet{devereux1998political} propose a model in which political instability causes governments to leave fewer assets to their successors which forces them into increasing capital taxes. The knowledge of future taxation then causes the private sector to reduce current investment which reduces future output.} and that policy uncertainty depresses business activity \citep{baker2016measuring}. Such sentiments are supported by empirical research that suggests political instability---as measured by regime change frequency or political violence---is negatively correlated with investment, financial development, and GDP growth in cross country regressions \citep{aisen2013does, alesina1996income, alesina1996political, fosu1992political, jong2009measurement, roe2011political} and that the variance of stock market returns increases in response to economic policy uncertainty \citep{leblang2005government, jensen2005market, liu2015economic}.

But such hypotheses are often overly simplistic. Not all political instability is equivalent - resignations, assassinations, and coups are all examples of regime changes, but may have disparate effects on financial returns. Resignations typically occur when an unpopular leader steps down and may signal the coming of a more effective leader. Assassinations are not always related to the effectiveness of a leader, can occur seemingly at random, and the successor to the assassinated leader may be unclear. Coups can occur in the name of democracy or autocracy, and in general, signal relative weakness in the current political system. Further, political instability may be seen as a positive event if the current regime is regarded as strongly anti-business or anti-global.  \citet{alesina1996political} take the possibility of a new government adopting better economic policies seriously, but argue that the negative effects of uncertainty dominate the positive effects of coups staged by pro growth factions. Likewise, \citet{londregan1990poverty} find that coups have no impact on economic growth and hypothesize that this may because there is a bimodal distribution of coups, some of which enhance growth and others which restrict it. More recent studies provide support for this hypothesis. \citet{meyersson2016political} finds that coup attempts that occur in democratic regimes are associated with negative GDP growth, while those that occur in autocratic regimes are associated with positive GDP growth. \citet{girardi2018institution} show that Salvador Allende's socialist government was associated with large declines in stock prices while the coup that replaced him boosted them.

In this study, we examine the impact of ``irregular'' regime changes and public protests on economic performance and remain agnostic about the direction of the effect. In contrast to the aforementioned cross country studies---but like \citet{girardi2018institution}---we conduct our analysis using daily financial data. Market expectations in the form of stock market returns are used as an indicator of whether the market views different types of potentially destabilizing political events as ``good'' for economic growth. The advantage of this event study approach is that it allows us to use time-series variation in stock returns to estimate treatment effects and mitigate the endogeneity problems that are nearly ubiquitous in cross country regressions.\footnote{A growing body of work uses event studies to assess political phenomenon.  For example, political events have been used to estimate the effect of political connections on firm value \citep[e.g.][]{fisman2001estimating,faccio2006politically,goldman2009politically}. Studies in ``forensic economics" have used abnormal returns from political events to locate or examine transactions such as insider trading \citep{dube2011coups}, illegal arms trading \citep{dellavigna2010detecting}, and the impact of hostilities on the financial performance of diamond mining firms in Angola \citep{guidolin2007diamonds}.} We analyze 13 successful coups, 8 assassinations, 15 forced resignations, and 11 public protests. We also examine the ultimately failed 2002 coup in Venezuela, as it provides a case that allows us to estimate the effects of both a seemingly successful pro-business coup, and the immediate reinstatement of a left-wing populist. Our study is therefore not only able to use an event-study approach to enhance the reliability of the estimates, but is not limited to a single case study as in \citet{girardi2018institution}. The relatively large sample enhances external validity and allows us to show that different types of political events have disparate economic effects.

@@

These findings are consistent with research that shows that ``good coups''---coups that lead to democratization or economic liberalization---are not the norm and therefore on average coups should not be expected to lead to increased economic development and positive market returns \citep{derpanopoulos2015coups, marinov2014coups, powell2011global, thyne2016coup, varol2011democratic}. However, while ``good coups'' may be the minority, regime changes that attempt to overthrow anti-business autocrats may be expected to result in positive abnormal returns as investors see little risk of a replacement government ``worse'' than the status quo.

@@

\section{Data}
Political data are primarily drawn from the Center for Systemic Peace's (CSP) Polity IV Coup d'etat dataset and Coup d'etat Events handbook. The Coup d'etat dataset includes the date of 1) successful coups, 2) attempted coups, 3) plotted coups and 4) alleged coup plots. We focus on successful coups because it is difficult to classify failed coups.\footnote{\citet[p. 617]{needler1966political} has even gone so far as to say that ``the categories of coups that were aborted, suppressed, or abandoned melt into each other and into a host of other non-coup phenomena so as to defy accounting.''} The one exception is that we separately study the failed coup attempt in Venezuela in April 2002 in which the President of Venezuela, Hugo Chavez, was removed from office for two days because it provides a natural test of the impact of the seemingly successful removal of a left-wing populist with a pro-business regime, and the ensuing reinstatement of a left-wing populist.

@@

The Coup d'etat Events handbook also provides a list of 1) auto-coups\footnote{Defined by Center for Systemic Peace as the ``occurrence of subversion of the constitutional order by a ruling (usually elected) executive and the imposition of an autocratic regime."}, 2) the ouster of leadership by foreign forces, 3) the ouster of leadership by rebel forces, 4) assassinations of the executive and 5) resignations of the executive due to poor performance and/or loss of authority. Daily financial data is available for countries in categories 4 and 5, so we supplement the coups with assassinations and resignations to form a list of ``irregular'' regime changes. The resignations are those in which the ruling executive was coerced to resign due to poor performance, public discontent and popular demonstrations.

@@

Financial data are from the Global Financial Data database, which includes the longest available time series of stock prices. We collect data on national equity indices and two global equity indices, the S\&P/IFC Emerging Market Investable Composite and the Morgan Stanley Capital International (MSCI) World Index. The S\&P/IFC index includes securities from emerging markets while the MSCI index includes securities from developed markets only. We collect stock index data on every country in which there was a coup or coup attempt, an assassination or failed assassination, or a forced resignation.\footnote{The list of failed assassinations are from \citet{jones2009hit}. Coup attempts are those in category 2 in the CSP Coup d'etat dataset.} The longest available daily time series for these stock indices are listed in \autoref{tab:stock-list}.

@@

Event day ARs are negative for all coups except the 1971 coup in Argentina, the 2002 coup in Nepal, and the 2002 failed coup d'etat attempt against Hugo Chavez (\autoref{fig:AR-Ven}). These results provide evidence that coups can in fact result in positive abnormal returns. While the 1971 Argentinian coup did result in another military leader, it did so while calling for free and democratic elections and replaced a government that had adopted extreme protectionist economic policies. In fact, by 1973 Argentina had transitioned to a democracy.\footnote{Based on Center for System Peace Polity IV polity score of 6. Values of 6-10 are defined as democracies.} The 2002 coup in Nepal resulted in a monarchical restoration, but occurred after the country's prime minister postponed general elections. The ultimately failed Venezuelan coup against Hugo Chavez replaced a left-wing populist government with a new pro-business president.

@@

The failed coup attempt against Hugo Chavez provides a natural test of the effects of both pro-business and anti-business regime changes separately from simple uncertainty because investors reacted to an expected regime change twice: first, when Chavez was ousted, and second, when he was reinstated. On the evening of April 11, 2002, coup plotters removed Chavez from office and later detained him. Pedro Carmona, a Venezuelan economist and business leader, was named the transitional President of Venezuela. Two days later, on April 13, 2002, a popular uprising led to Chavez's reinstatement as president. This provides an estimate of the market's valuation of a transition from the Chavez regime to the Carmona regime and its valuation of a transition from the Carmona regime back to the Chavez regime. By extension, it provides an estimate of the impact of a shift from a left-wing populist government to a pro business regime in an emerging market.

\autoref{fig:AR-Ven} provides graphical evidence on the effect of the coup attempt. The top panel shows CARs for the 10 days prior to and following the event, along with 95\% confidence intervals. The daily ARs and corresponding confidence intervals are displayed in the bottom panel. The abnormal return on April 12, the first trading day in which investors could react to the coup, was 10\%. The market reacted in the opposite direction to Chavez's reinstatement as president: the abnormal return on the next trading day, April 15, 2002, was -8\%.

@@

>>>>

The image of a coup or popular uprising ousting an autocratic ruler and installing a democratic government is an evocative one. However, both recent events and prior research suggest that coups rarely lead to democracy. In the wake of the Arab Spring---in which rulers were deposed in four countries---only Tunisia ultimately transitioned to democracy. These results are in line with prior research that claims that in autocracies coups rarely lead to democratization, and in democracies they cause instability \citep{derpanopoulos2015coups, powell2011global, thyne2016coup, varol2011democratic}.

Of course, coups and other regime changes not only impact a country's political environment but its economy as well. For instance, investors, multinational firms, development agencies, and aid organizations rank political risk as a top consideration when making investment decisions in emerging markets.\footnote{In the 2013 Multilateral Investment Guarantee Agency (MIGA) \citet{wipr2013} report (the last report published), executives of multinational enterprises (MNEs) ranked political risk as the second most important constraint for foreign direct investment (FDI) in developing countries over the next three years (after macroeconomic instability). In \citet{wipr2011} and  \citet{wipr2012}, political risk was ranked the most important constraint for FDI in developing countries - even greater than macroeconomic instability. The types of political risk of most concern to investors in developing countries (ranked in order of importance) were 1) adverse regulatory changes, 2) breaches of contract, 3) transfers and convertibility restrictions, 4) civil disturbances, 5) non-honoring of government guarantees, 6) expropriation nationalization, 7) terrorism and 8) war.} Common rationales are that unstable governments act myopically and adopt inefficient policies. Unstable governments have, for example, been argued to be less likely to invest in the legal system and the protection of property rights \citep{svensson1998investment}, and to be more likely to increases taxes \citep{devereux1998political}.\footnote{\citet{devereux1998political} proposes a model in which political instability causes governments to leave fewer assets to their successors which forces them into increasing capital taxes. The knowledge of future taxation then causes the private sector to reduce current investment which reduces future output.} Such sentiments are supported by empirical research that suggests political instability---as measured by regime change frequency, political violence, or indices of uncertainty derived from news reports---is negatively correlated with investment, financial development, and GDP growth in cross country regressions \citep{aisen2013does, alesina1996income, alesina1996political, fosu1992political, jong2009measurement, roe2011political, baker2016measuring} and that the variance of stock market returns increases in response to economic policy uncertainty \citep{leblang2005government, jensen2005market, liu2015economic}.

But such hypotheses may be overly simplistic. Not all political instability is equivalent--- resignations, assassinations, and coups are all examples of regime changes, but may have disparate effects on financial returns. Resignations typically occur when an unpopular leader steps down and may signal the coming of a more effective leader. Assassinations are not always related to the effectiveness of a leader, can occur seemingly at random, and the successor to the assassinated leader may be unclear. Coups can occur in the name of democracy or autocracy, and in general, signal relative weakness in the current political system. In other words, investors may view resignations as more stable than coups or assassinations, despite also representing a change in regime. Further, political instability may actually be seen as a positive event if the current regime is regarded as strongly anti-business or anti-global.  \citet{alesina1996political} take the possibility of a new government adopting better economic policies seriously, but argue that the negative effects of uncertainty dominate the positive effects of coups staged by pro growth factions. Likewise, \citet{londregan1990poverty} find that coups have no impact on economic growth and hypothesize that this may because there is a bimodal distribution of coups, some of which enhance growth and others which restrict it. More recent studies provide support for this hypothesis. \citet{meyersson2016political} finds that coup attempts that occur in democratic regimes are associated with negative GDP growth, while those that occur in autocratic regimes are associated with positive GDP growth. \citet{girardi2018institution} show that Salvador Allende's socialist government was associated with large declines in stock prices while the coup that replaced him boosted them.

In this study, we examine the impact of ``irregular'' regime changes and public protests on economic performance and remain agnostic about the direction of the effect. In contrast to the aforementioned cross country studies---but like \citet{girardi2018institution}---we conduct our analysis using daily financial data. Market expectations in the form of stock market returns are used as an indicator of whether the market views different types of potentially destabilizing political events as ``good'' for economic growth. The advantage of this event study approach is that it allows us to use time-series variation in stock returns to estimate treatment effects and mitigate the endogeneity problems that are nearly ubiquitous in cross country regressions.\footnote{A growing body of work uses event studies to assess political phenomenon.  For example, political events have been used to estimate the effect of political connections on firm value \citep[e.g.][]{fisman2001estimating,faccio2006politically,goldman2009politically}. Studies in ``forensic economics" have used abnormal returns from political events to locate or examine transactions such as insider trading \citep{dube2011coups}, illegal arms trading \citep{dellavigna2010detecting}, and the impact of hostilities on the financial performance of diamond mining firms in Angola \citep{guidolin2007diamonds}.} We analyze 13 successful coups, 8 assassinations, 15 forced resignations, and 11 public protests. We recognize that this does not represent a full list of historical coups, assassinations, resignations, and public protests, but are restricted to the sample of events for which national-level daily financial data is available. We also examine the ultimately failed 2002 coup in Venezuela, as it provides a case that allows us to estimate the effects of both a seemingly successful pro-business coup, and the immediate reinstatement of a left-wing populist. Our study is therefore not only able to use an event-study approach to enhance the reliability of the estimates, but is not limited to a single case study as in \citet{girardi2018institution}. The relatively large sample enhances external validity and allows us to show that different types of political events have disparate economic effects.

@@

These findings are consistent with research that shows that ``good coups''---coups that lead to democratization or economic liberalization---are not the norm \citep{derpanopoulos2015coups, powell2011global, thyne2016coup, varol2011democratic}. However, while ``good coups'' may be the minority, some evidence suggests that they have become more frequent \citep{marinov2014coups}, and regime changes that attempt to overthrow anti-business autocrats may be expected to result in positive abnormal returns as investors see little risk of a replacement government ``worse'' than the status quo. Previous research therefore suggests that on average coups should not be expected to lead to increased economic development and positive market returns, but that positive returns associated with ``good coups'' should also occasionally be present. We provide the first systematic empirical evidence in support of this theory.

@@

\section{Data}

Financial data are from the Global Financial Data database, which includes the longest available time series of stock prices. We collect data on national equity indices and two global equity indices, the S\&P/IFC Emerging Market Investable Composite and the Morgan Stanley Capital International (MSCI) World Index. The S\&P/IFC index includes securities from emerging markets while the MSCI index includes securities from developed markets only. We collect national stock index data on every country in which there was a coup or coup attempt, an assassination or failed assassination, or a forced resignation and for which daily financial data is available.\footnote{The list of failed assassinations are from \citet{jones2009hit}. Coup attempts are those in category 2 in the CSP Coup d'etat dataset.} The longest available daily time series for these stock indices are listed in \autoref{tab:stock-list}.

Political data are primarily drawn from the Center for Systemic Peace's (CSP) Polity IV Coup d'etat dataset and Coup d'etat Events handbook. The Coup d'etat dataset includes the date of 1) successful coups, 2) attempted coups, 3) plotted coups and 4) alleged coup plots. We focus on successful coups because it is difficult to classify failed coups. \citet[p. 617]{needler1966political} has even gone so far as to say that ``the categories of coups that were aborted, suppressed, or abandoned melt into each other and into a host of other non-coup phenomena so as to defy accounting.'' The one exception is that we separately study the failed coup attempt in Venezuela in April 2002 in which the President of Venezuela, Hugo Chavez, was removed from office for two days because it provides a natural test of the impact of the seemingly successful removal of a left-wing populist with a pro-business regime, and the ensuing reinstatement of a left-wing populist.

@@

The Coup d'etat Events handbook also provides a list of 1) auto-coups\footnote{Defined by Center for Systemic Peace as the ``occurrence of subversion of the constitutional order by a ruling (usually elected) executive and the imposition of an autocratic regime."}, 2) the ouster of leadership by foreign forces, 3) the ouster of leadership by rebel forces, 4) assassinations of the executive and 5) resignations of the executive due to poor performance and/or loss of authority. Daily financial data is available for countries in categories 4 and 5, so we supplement the coups with assassinations and resignations to form a list of ``irregular'' regime changes. The resignations are those in which the ruling executive was coerced to resign due to poor performance, public discontent and popular demonstrations. Note that the Polity IV definition of ``poor performance'' is not synonymous with poor \textit{economic} performance, and in practice the reasons cited for resignation across events are: loss in conflict/war, anti-authoritarian protest, corruption scandals, Supreme Court ruling against unconstitutional actions, contested elections, and abuse of power.

%loss in conflict/war (Argentina), anti-authoritarian protest (Philippines, Bangladesh, Nepal, Tunisia), corruption scandal (Philippines, Lithuania), constitutional ruling (Pakistan), contested election (Ukraine), abuse of power (Ecuador) Thailand?

@@

@@

Event day ARs are negative for all coups except the 1971 coup in Argentina, the 2002 coup in Nepal, and the 2002 failed coup d'etat attempt against Hugo Chavez (\autoref{fig:AR-Ven}). These results provide evidence that coups do not necessarily lead to negative abnormal returns. While the 1971 Argentinian coup did result in another military leader, it did so while calling for free and democratic elections and replaced a government that had adopted extreme protectionist economic policies. In fact, by 1973 Argentina had transitioned to a democracy.\footnote{Based on Center for System Peace Polity IV polity score of 6. Values of 6-10 are defined as democracies.} The 2002 coup in Nepal resulted in a monarchical restoration, but occurred after the country's prime minister postponed general elections. The ultimately failed Venezuelan coup against Hugo Chavez replaced a left-wing populist government with a new pro-business president.

@@

The failed coup attempt against Hugo Chavez therefore provides a natural test of the effects of both pro-business and anti-business regime changes separately from simple uncertainty because investors reacted to an expected regime change twice: first, when Chavez was ousted, and second, when he was reinstated. On the evening of April 11, 2002, coup plotters removed Chavez from office and later detained him. Pedro Carmona, a Venezuelan economist and business leader, was named the transitional President of Venezuela. Two days later, on April 13, 2002, a popular uprising led to Chavez's reinstatement as president. This provides an estimate of the market's valuation of a transition from the Chavez regime to the Carmona regime and its valuation of a transition from the Carmona regime back to the Chavez regime. By extension, it provides an estimate of the impact of a shift from a left-wing populist government to a pro business regime in an emerging market.

\autoref{fig:AR-Ven} provides graphical evidence on the effect of the coup attempt. The top panel shows CARs for the 10 days prior to and following the event, along with 95\% confidence intervals. The daily ARs and corresponding confidence intervals are displayed in the bottom panel. The abnormal return on April 12, the first trading day in which investors could react to the coup, was 10\%. The market reacted in the opposite direction to Chavez's reinstatement as president: the abnormal return on the next trading day, April 15, 2002, was -8\%. We also note that Chavez's own failed attempt to overthrow the democratically elected government of Venezuela--- at the time pursuing deficit reduction efforts in order to obtain IMF loans---in 1992 was similarly viewed negatively by investors, resulting in an approximately -6\% AR (see figure \autoref{fig:AR-Ven-1992}). These events therefore demonstrate a positive market reaction to the attempted overthrow of a socialist leader, as well as two examples of negative reactions to attempted and successful overthrows of pro-market leaders.

@@

\subsection{Additional graphical depictions of individual events}

\begin{figure}[!htb]
\includegraphics{../figs/venezuela_coup_attempt_1992.pdf}
\caption{Abnormal returns surrounding the 1992 Venezuelan coup d'etat attempt}
\label{fig:AR-Ven-1992}
\end{figure}
